---
title: "State variability in racial stereotypes"
output: 
  html_document:
    toc: true
---
  
# Description of Data
This script performs a state-level regression predicting state-level variation in the stereotype that Black Americans are violent as a function of state-level crime rate.
  
# Set-up environment

## Import relevant libraries
```{r installPackages, message = FALSE}
library(ggplot2) # For general plotting
library(usmap) # For making plots of US
library(egg) # For multiple plots
library(dplyr) # For data manipulation (part of tidyverse)
library(tidyr) # For making tidy data (part of tidyverse)
```

## Import and clean data from the Uniform Crime Report (UCR)
```{r loadData, results = "hide"}
#Load 2006-2014 crime data Uniform Crime Report
ucr <- read.csv("UCR06-14.csv")

#Rename variables for easier coding
names(ucr) <- c("state", "year", "offense", "white", "black")

#Change offense variable into labeled factor 
ucr$offense <- factor(ucr$offense, labels = c("murder", "assault"))

#Recode state names, note FL doesn't provide data
ucr$state <- as.numeric(gsub("[^[:digit:].]", "",  ucr$state)) #Regex to extract number
ucr$stateAbb <- recode(ucr$state, '1' = "AL", '2' = "AZ", '3' = "AR", '4' = "CA", 
  '5' = "CO", '6' = "CT", '7' = "DE", '8' = "DC", '9' = "FL", '10' = "GA", '11' = "ID", 
  '12' = "IL", '13' = "IN", '14' = "IA", '15' = "KS", '16' = "KY", '17' = "LA", 
  '18' = "ME", '19' = "MD", '20' = "MA", '21' = "MI", '22' = "MN", '23' = "MS", 
  '24' = "MO", '25' = "MT", '26' = "NE", '27' = "NV", '28' = "NH", '29' = "NJ", 
  '30' = "NM", '31' = "NY", '32' = "NC", '33' = "ND", '34' = "OH", '35' = "OK", 
  '36' = "OR", '37' = "PA", '38' = "RI", '39' = "SC", '40' = "SD", '41' = "TN",
  '42' = "TX", '43' = "UT", '44' = "VT", '45' = "VA", '46' = "WA", '47' = "WV", 
  '48' = "WI", '49' = "WY", '50' = "AK", '51' = "HI") #state abbreviations
```

## Aggregate data to the state level
```{r aggState, results = "hide"}
#Pivot the data frame so that each line is one state
ucr <- as.data.frame(ucr %>% #Using ucr data
  group_by(stateAbb, offense) %>% #group by state and offense
  summarize(W = sum(white), B = sum(black)) %>% #collapse across years
  pivot_wider( #function to pivot the data
    names_from =  c(offense), #crime type as column
    values_from = c(W, B), #values from crime counts
    names_sep = "") %>%
  rename(murderW = Wmurder, assaultW = Wassault,
    murderB = Bmurder, assaultB = Bassault))
```

## Load homicide and census data and merge
```{r loadState, results = "hide"}
#Import other state data, already aggregated for the sake of the example 
state <- read.csv("stateData.csv")

#Merge state-level data with crime data aggregated to the state level
df <- merge(state, ucr, by = "stateAbb", all = T)
```


## Feature Manipulation
```{r featureManip, message = FALSE}
df$deathRateB <- round((df$deathsB / df$populationB)*100000, 3)
df$assaultRateB <- round((df$assaultB / df$populationB)*100000, 3)
df$murderRateB <- round((df$murderB / df$populationB)*100000, 3)
df$deathRateW <- round((df$deathsW / df$populationW)*100000, 3)
df$assaultRateW <- round((df$assaultW / df$populationW)*100000, 3)
df$murderRateW <- round((df$murderW / df$populationW)*100000, 3)
df$sexRatio <- round(df$male / df$female, 4)
df$perWhite <- round(df$white / df$population2010, 4)

#Convert state code to character for mapping
df$fips <- as.character(df$fips)
df$fips[1:7] <- c("01", "02", "04", "05", "06", "08", "09")
```

# Descriptives
Survey asked whether respondents associated Whites or Blacks more with violence

On a 1-7 scale, higher values indicate Blacks associated with violence

1 = Whites violent, 4 = Neutral, 7 = Blacks violent

```{r descriptives, message = FALSE}
df[order(df$blackViolent, decreasing = T)[1:5], c("state", "blackViolent")] #highest
df[order(df$blackViolent, decreasing = F)[1:5], c("state", "blackViolent")] #lowest
```

## Map of state variation
```{r stateMap, message = FALSE}
#Plot State variation in the Black violence stereotype
expMap <- plot_usmap(regions = "states", #plot US states
    data = df, #data frame is df
    values = "blackViolent") + #Plot black-violence stereotype
  ggtitle("Black Violence Stereotype by State") + 
  scale_fill_continuous(low = "white", high = "dark red", limits = c(4.0, 4.5),
    guide = "colorbar", name = "Stereotype") + 
  theme(legend.position = "right", #"right"
    legend.title = element_blank(), 
    plot.title = element_text(hjust = .5))

#Display plot
expMap
```

# Regression Models

Predictive model of state-level prejudice in the stereotype that Blacks are violent

blackViolent measures how much Black (vs White) Americans associated with violence

```{r regMurder, message = FALSE}

contMod <- lm(scale(blackViolent) ~ scale(sexRatio) + scale(age) + scale(perWhite) + 
  scale(income) + scale(bachelors) + scale(con), df) #control variables

#Update model to add in key predictors (race-specific crime rate)
fullMod <- update(contMod, ~ . + scale(murderRateB) + scale(murderRateW))

#Examine model coefficients
summary(fullMod)

#Assess prediction quality
round(summary(fullMod)$r.squared, 2) #Full model explains 58% of variance
round(summary(fullMod)$r.squared - summary(contMod)$r.squared, 2) #9% more than control  
```

## Results with murder data
 
In states with higher murder rates among White Americans, the Black-violent stereotype is weaker.

In states with higher murder rates among Black Americans, the Black-violent stereotype is stronger.

## Replication with homicide data

Repeat model above but using a different proxy for crime

Homicide rates as determined by CDC data

Update model to add in key predictors (race-specific crime rate)

```{r regHomicide, message = FALSE}
fullMod2 <- update(contMod, ~ . + scale(deathRateB) + scale(deathRateW))

summary(fullMod2)

#Assess prediction quality
round(summary(fullMod2)$r.squared, 2) #Full model explains 73% of variance
round(summary(fullMod2)$r.squared - summary(contMod)$r.squared, 2) #24% more than control  
```

## Results with homicide data

In states with higher homicide rates among White Americans, the Black-violent stereotype is weaker.

In states with higher homicide rates among Black Americans, the Black-violent stereotype is stronger.

## Plots of state correlations

```{r statePlots, message = FALSE, warning = FALSE}
deathW <- ggplot(df, aes(y = blackViolent, x = murderRateW, label = stateAbb)) + 
  geom_text(aes(label = stateAbb), hjust = -.1, vjust = 0, size = 1.5) + 
  geom_abline(intercept = coef(lm(blackViolent ~ deathRateW, data = df))[[1]], 
    slope = coef(lm(blackViolent ~ deathRateW, data = df))[[2]]) +
  ylab("Black-Violence Stereotype") + 
  xlab("White Death Rate") + 
  theme_bw(base_size = 10) + 
  theme(
    axis.title.y = element_text(hjust = .5),
    axis.title.x = element_text(hjust = .5),
    panel.background = element_blank(), 
    panel.grid = element_blank(),
    panel.border = element_rect(colour = "black", fill = NA, size = 1))

deathB <- ggplot(df, aes(y = blackViolent, x = deathRateB, label = stateAbb)) + 
  geom_text(aes(label = stateAbb), hjust = -.1, vjust = 0, size = 1.5) + 
  geom_abline(intercept = coef(lm(blackViolent ~ murderRateB, data = df))[[1]], 
    slope = coef(lm(blackViolent ~ deathRateB, data = df))[[2]]) +
  ylab("Black-Violence Stereotype") + 
  xlab("Black Death Rate") + 
  theme_bw(base_size = 10) + 
  theme(
    axis.title.y = element_text(hjust = .5),
    axis.title.x = element_text(hjust = .5),
    panel.background = element_blank(), 
    panel.grid = element_blank(),
    panel.border = element_rect(colour = "black", fill = NA, size = 1))

murderW <- ggplot(df, aes(y = blackViolent, x = murderRateW, label = stateAbb)) + 
  geom_text(aes(label = stateAbb), hjust = -.1, vjust = 0, size = 1.5) + 
  geom_abline(intercept = coef(lm(blackViolent ~ murderRateW, data = df))[[1]], 
    slope = coef(lm(blackViolent ~ murderRateW, data = df))[[2]]) +
  ylab("Black-Violence Stereotype") + 
  xlab("White Murder Rate") + 
  theme_bw(base_size = 10) + 
  theme(
    axis.title.y = element_text(hjust = .5),
    axis.title.x = element_text(hjust = .5),
    panel.background = element_blank(), 
    panel.grid = element_blank(),
    panel.border = element_rect(colour = "black", fill = NA, size = 1))

murderB <- ggplot(df, aes(y = blackViolent, x = murderRateB, label = stateAbb)) + 
  geom_text(aes(label = stateAbb), hjust = -.1, vjust = 0, size = 1.5) + 
  geom_abline(intercept = coef(lm(blackViolent ~ murderRateB, data = df))[[1]], 
    slope = coef(lm(blackViolent ~ murderRateB, data = df))[[2]]) +
  ylab("Black-Violence Stereotype") + 
  xlab("Black Murder Rate") + 
  theme_bw(base_size = 10) + 
  theme(
    axis.title.y = element_text(hjust = .5),
    axis.title.x = element_text(hjust = .5),
    panel.background = element_blank(), 
    panel.grid = element_blank(),
    panel.border = element_rect(colour = "black", fill = NA, size = 1))

#Multiple plots
#Missing rows due to FL murder data not reported
ggarrange(deathW, deathB, murderW, murderB, nrow = 2, ncol = 2)
```





